Neural Connectome of the Ctenophore Statocyst

### 2025-07-10 manuscript/Manuscript.qmd
Ms: reorganised sections in introduction

To address the lack of structural, circuit-level understanding of the ctenophore nervous system, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst formed by lithocytes and four clusters of mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980]. In the natural environment, most ctenophores can freely adjust their posture within the water column and exhibit both positively and negatively gravitactic orientations. This behavioral flexibility is likely supported by the aboral organ, which functions as a gravity sensor and pacemaker for ciliary movement. In this study, we focused on the orientation in which the aboral organ faces upward ("balancer-up" posture), as this configuration facilitates intuitive interpretation of balance-like functions and matches the setup used in high-speed imaging experiments. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011].

A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed-direction response as seen in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that ionic currents and potential neural inputs play critical roles in shaping balancer-cell output [@Lowe1997].

To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*. Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cells. By functional imaging, we uncovered correlated activity patterns in balancer cilia across the four quandrants of the apical organ. Our results suggest that the aboral nerve net functions in coordinating the activity of balancer cilia, rather than fulfilling a feedforward sensory-motor function.

>>>>

To address the lack of structural, circuit-level understanding of the ctenophore nervous system, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst formed by lithocytes and four clusters of mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980]. 
A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. 
Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system [@HORRIDGE1965; @Tamm1982; @tamm2014b].

In the natural environment, most ctenophores can freely adjust their posture within the water column and exhibit both positively and negatively gravitactic orientations. This behavioral flexibility is likely supported by the aboral organ, which functions as a gravity sensor and pacemaker for ciliary movement. In this study, we focused on the orientation in which the aboral organ faces upward ("balancer-up" posture), as this configuration facilitates intuitive interpretation of balance-like functions and matches the setup used in high-speed imaging experiments.

Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating the balancer response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. 
Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. 
These findings support the hypothesis that ionic currents and potential neural inputs play critical roles in shaping balancer-cell output [@Lowe1997].

Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011]. However, how this deep nerve net regulates balancer or other functions has remained elusive.

To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy (vEM) and high-speed imaging in the ctenophore *Mnemiopsis leidyi*. 
Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cells. 
By functional imaging, we uncovered correlated activity patterns in balancer cilia across the four quandrants of the apical organ.
Our results suggest that the aboral nerve net functions in coordinating the activity of balancer cilia, rather than fulfilling a feedforward sensory-motor function.
